---
title: "HiCaptuRe Functions"
author: "Laureano Tomás-Daza"
date: "2023-12-11"
output: 
  html_document:
    css: style.css
    theme: simplex
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Functions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this vignette we will show you the main functions of HiCapture and
how they work. To start playing with the function first we need some
example data provided in the package. Let's save them in variables:

```{r, example_data}
ibed1 <- system.file("extdata", "ibed1_example.ibed", package="HiCaptuRe")
ibed2 <- system.file("extdata", "ibed2_example.ibed", package="HiCaptuRe")
annotation <- system.file("extdata", "annotation_example.txt", package="HiCaptuRe")
regions <- system.file("extdata", "regions_example.bed", package="HiCaptuRe")
```

```{r, comment=F, message=F,library}
library(HiCaptuRe)
```

# Importing interaction data: `load_interactions`

The first step is loading the interaction file into R, to do so we use
the function `load_interactions`.

This function reads the file and removes some "technical artefacts" of
Chicago, like duplicated interactions. And it gives us the interactions
in an easy readable format, a `HiCaptuRe` object, which contains a
`GenomicInteractions` object.

We can load different types of interaction files: ibed, seqmonk, washU
(and it old format), bedpe and peakmatrix. The function detect
automatically which is the format of our file.

```{r,load_interactions}
ibed1 <- load_interactions(file = ibed1)
```

We can see that it detects that the file is in ibed format. And the
function removes several interactions duplicated.

```{r,ibed1}
ibed1
```

The resulting `HiCaptuRe` object looks like a `GenomicInteractions`
object but with some extra slots. This object contains the coordinates
of both anchors of the interaction and metadata:

\- `bait_1` and `bait_2` : annotation of each restriction fragment, in
case one of them does not have annotation it shows a dot (".")

\- `ID_1` and `ID_2` : the ID of the restriction fragment when digesting
the genome with an specific restriction enzyme. This ID comes from the
function `digest_genome` .

\- `reads` : The number of reads detected in each interaction.

\- `CS` : The Chicago Score of each interaction.

\- `counts` : It's a column added when creating a `GenomicInteractions`
object, that tells the number of times that each interaction appears, in
this case it will always be 1.

\- `int` : The type of interaction. It could be B_B or B_OE, meaning
interaction between 2 baits (B_B) or between a bait and an other-end
(B_OE).

\- `distance` : The distance between the midpoint of the anchors of the
interactions.

**NOTE**: When loading certain type of files, such as WashU format or
bedpe, they lack of some information such as the number of reads or the
annotation. In these cases the function assigns a value to these
columns. In the annotation it adds "non-annotated" to all fragments and
in the reads columns it adds a zero to all of them.

Now, let's have a look at the new extra slots of the `HiCaptuRe` object.

```{r,slots}
slotNames(ibed1)
```

There are several slots that belong to the `GenomicInteractions` object,
the new ones are: **parameters**, **ByBaits** and **ByRegions**.

These slots are filled while running certain functions, with this
function we have only initialize the slot parameters, so let's have a
look at it:

```{r, parameters}
ibed1@parameters
```

This slot contains a list with the parameters used to run each function.

So we can see that the last step that we run was "load" and the 2
arguments used in this function. But we can see that there is another
step before called "digest". This step refers to the function
`digest_genome`.

## Digesting the Genome: `digest_genome`

This function generates the restriction fragments resulting from the
virtual digestion of a given genome. The default arguments are for the
restriction enzyme HindIII in the genome version GRCh38.

There are some interesting arguments:

-   `RE_name`, `motif` and `cut_position`. These arguments refer to the
    restriction enzyme used to digest the genome. We have to indicate
    the recognition motif of the enzyme and the cut position of the
    enzyme inside that motif.

<!-- -->

-   `select_chr`. In most genome there are more chromosomes than the
    commonly known (1:22, X, Y), so here we can specify which
    chromosomes we want to digest

-   `PAR_mask` and `PAR_file`. Pseudoautosomical Regions (PAR) are
    regions of the sexual chromosomes that are equal between X and Y,
    and in certain genomes might be masked. For example in Ensembl human
    genome the PAR region of chromosome Y is mask, while in UCSC version
    is not. This fact affects the digestion of the genome, only in
    chromosome Y, generating different restriction fragments depending
    on the version. With these 2 arguments we can specify if we want to
    mask those regions and provide a file with the coordinates of such
    regions. The package already contains the PAR coordinates for the
    GRCh38 genome.

**NOTE**: This function is internally called each time we run
`load_interactions` to provide the ID of the restriction fragment so
each file that we load have the same IDs. The first time that
`digest_genome` is called takes a while to digest a certain genome, but
the next times is much faster since the digested genome is saved on
cache.

# Annotating interaction object: `annotate_interactions`

This function allows us to annotate the baits of the interactions,
particularly useful when loading files without annotation such as WashU
format or bedpe files.

When performing a C-HiC experiment we need a capture library, that
essentially capture certain regions of the genome of our interest. The
annotation that we used here is that capture library, meaning, for each
restriction fragment of interest we need an annotation (gene name,
enhancer ID, copy number variant ID, etc.).

```{r, annotate}
ibed_annotated <- annotate_interactions(interactions = ibed1,
                                        annotation = annotation)
ibed_annotated
```

We can see that in the object `ibed1` the columns bait_1 and bait_2
contains Ensembl transcript ID, while now it contains a mix of Ensembl
gene ID and gene names, because we change its annotation.

As we saw before the slot **parameters** is updated again with the
arguments used in this function.

```{r, annotate_parameters}
ibed_annotated@parameters
```

# Filtering interactions by Baits of interest: `interactionsByBaits`

This function subset the interactions by any given bait or list of
baits, meaning, it returns the interactions which any anchor of the
interaction contains a bait of interest.

Since out data is focused on promoters, we might be interested in
subsetting our data to only select those interactions that involved a
set of promoters of interest. These promoters of interest could came
from an RNA-seq analysis, or any other type of analysis.

For example, we might be interested in these 3 genes, so we used the
function to subset the interactions.

```{r, interactionsByBaits}
baits_of_interest <- c("MTOR","APOA2","TP53")
ibed_byBaits <- interactionsByBaits(interactions = ibed_annotated,
                                    baits = baits_of_interest)
```

The function informs us that some baits do not have any interactions.

```{r,interactionsByBaits_show}
ibed_byBaits
```

Now we have a subset with only 18 interactions containing our baits of
interest.

As previous functions, the slop **parameters** is updated again, but now
we will check the slop **ByBaits**, that is only updated by this
function.

```{r, ByBaits}
ibed_byBaits@ByBaits
```

This slot gives us a **bait-centric** summary, so for each bait of
interests it gives us its fragment ID, the number of interactions in
which is involved, the number of other-end that are interacting with it,
the ID of the interacting regions and their annotations.

Each time we run this function a new element is added to this slot, so
we can keep track of the different times that we have run this function.

# Integrating Regions of interest: `interactionsByRegions`

This function subset the interactions by any given region or set of
regions, meaning, it returns the interactions which any anchor of the
interaction overlaps with a region of interest.

This function is particularly useful when we want to integrate regions
from other omic, such as ChIP-seq, CUT&RUN, copy number variants, etc.

```{r, interactionsByRegions}
ibed_byRegions <- interactionsByRegions(interactions = ibed_annotated,
                                        regions = regions)
```

The function informs us of the resulting number of interactions.

```{r, interactionsByRegions_show}
ibed_byRegions
```

This function adds 8 extra metadata columns, 4 for each anchor. These
columns are:

-   `region_*`: T/F column indicating if a region overlap this anchor.

-   `Nregion_*`: numeric column indicating the number of regions
    overlapping this anchor.

-   `regionID_*`: the ID of the regions overlapping the anchor.

-   `regionCov_*`: the number of base pairs of the anchor overlapping
    with the region. In case more than one region overlap it's the sum
    of the base pairs of all regions overlapping.

As previous functions, the slop **parameters** is updated again, but now
we will check the slop **ByRegions**, that is only updated by this
function.

```{r, byRegions}
ibed_byRegions@ByRegions
```

This slot gives us a **region-centric** summary, so for each region of
interests it gives us its ID, the number of fragments, the number of
Other-End fragments, and the fragment ID that overlaps.

Each time we run this function a new element is added to this slot, so
we can keep track of the different times that we have run this function.

# Intersecting sets of interactions: `intersect_interactions`

This function intersects the several interactions set to classify the
interactions depending on the samples where are present, basically a
classic Venn diagram.

To use this function we need at least 2 `HiCaptuRe` objects in a named
list, so we need to load another interaction file.

```{r, ibed2}
ibed2 <- load_interactions(file = ibed2)
ibed2_annotated <- annotate_interactions(interactions = ibed2, annotation = annotation)
```

```{r, intersect_interactions}
interactions_list <- list(A = ibed_annotated, B = ibed2_annotated)
output <- intersect_interactions(interactions_list = interactions_list)
```

This function gives us a list with 3 elements:

-   **intersections**: contains a list of `HiCaptuRe` objects of the
    resulting intersections. In this case the specific interactions of
    A, of B and the common ones.

```{r, intersections}
lapply(output$intersections, function(x) x[1:2])
```

-   **upset**: contains an upset plot of the intersections of
    interactions

```{r, upset, out.width="50%"}
output$upset
```

-   **venn**: upset: contains a venn diagram of the intersections of
    interactions

```{r, venn, out.width="50%"}
output$venn
```

# Summarise interactions by distance: `distance_summary`

This function allows us to get a summary of the number of interactions
by their distance. This is interesting to compare the profile of
distances between conditions.

```{r, distance_summary}
dist_sum <- distance_summary(interactions = ibed_annotated,
                             breaks = seq(0, 10^6, 10^5),
                             sample = "ibed1")
dist_sum
```

The output is a table with several columns that gives us information of
each type of interaction in each range of distance. The columns are:

-   **int**. Type of interaction: B_B, B_OE or Total (B_B + B_OE)

-   **total_per_int**. Total number of interaction of each type

-   **sample**. Name of the sample

-   **HiCaptuRe**. Total number of interactions present in the input
    `HiCaptuRe` object.

-   **break**. Distance range

-   **value**. Number of interactions of this type in this break.

# Plotting the summary of distance: `plot_distance_summary`

This function plots the output of the previous one. We can plot the
values in 3 different ways:

-   **Absolute**: the values without any modification.

```{r, absolute, fig.show="hold", out.width="50%"}
plots <- plot_distance_summary(distances = dist_sum, type_of_value = "absolute")
plots$int_dist
plots$total_dist
```

-   **by_int_type**: the values divided by the total number of
    interactions of each type.

```{r, by_int_type, out.width="50%"}
plots <- plot_distance_summary(distances = dist_sum, type_of_value = "by_int_type")
plots$int_dist_norm_int

```

-   **by_total**: the values divided by the total number of interactions
    in the `HiCaptuRe` object.

```{r, by_total, fig.show="hold",  out.width="50%"}
plots <- plot_distance_summary(distances = dist_sum, type_of_value = "by_total")
plots$int_dist_norm_total
plots$total_dist_norm_total
```

# Exporting proccessed interaction data: `export_interactions`

This is the final function. It used to export the interactions that we
have generated using all the previous functions.

The `HiCaptuRe` object can be exported in several different format,
being the default **ibed** format. The other formats are: peakmatrix,
washU, washUold, cytoscape and bedpe.

```{r, export, eval=F}
export_interactions(interactions = ibed_annotated,
                    file = "/path/to/folder/ibed_annotated.ibed",
                    type = "ibed")
```

**NOTE**: the peakmatrix format can only be used if the **HiCaptuRe**
object comes from a peakmatrix. And if you want to export a peakmatrix
in any other format several files will be exported, one per each sample
present in the peakmatrix.
